How to Avoid OOTA Without Really Trying

### 2024-10-19 paper/ootavacuous.tex
ootavacuous: More fairly minor edits
Some more edits to improve the flow of the text.  Only one of them, in
6.1, is crucial; it fixes a typo (a word that was left out).

Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.

This paper uses \emph{full C++} to denote the standard including this
prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
that excludes this prohibition but is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

In some cases, the standard permits architectures to differ as to whether
an execution contains a semantic dependency.
For example, consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
running on implementations in
which \co{char} is a signed type,
but not those for which it is unsigned.

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).\footnote{
	This requirement is the standard's ``as-if'' rule.}
This means:
(1)~The program's output must be the same as that of the abstract execution;
(2)~Volatile accesses ``are evaluated strictly according to the
rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
(3)~(A condition on the timing and interleaving of input and output,
which does not matter for our purposes).
We will say that the abstract execution is \emph{realized by} the
hardware execution.

@@

Conditional or computed branches also provide ordering.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional- or computed-branch
instruction must commit after the branch, and hence after the source
for the branch's condition and/or destination
input.\footnote{
	One exception is a branch that conditionally jumps to the
	next instruction.
	We ignore this possibility by treating such branches as no-ops.}

@@

We begin by recognizing that semantic dependencies are relative to a
particular execution and a particular implementation,
as described in Section~\ref{sec:Properties of Semantic Dependencies}.
The same source code may or may not contain a semantic dependency,
according to the details of the execution in question and the
machine code produced by the compiler.
For this reason we will characterize semantic dependencies in a given
abstract execution realized by a given hardware execution.
(While it possible to argue about semantic dependencies in abstract
executions that have no hardware realizations, doing so seems pointless.)

@@

Let $W'$ and $R'_i$ be the accesses in $H$ that $W$ and the
non-omitted $R_i$ loads map to.
We then require that the hardware executions of $T$ in $H$ and $G$ be
identical for an initial period lasting up to the first of the $R'_i$.
Following this initial period there will be a common period, during
which $H$ and $G$ execute the same machine instructions but do not
necessarily compute the same values.
This common period ends when one of the hardware executions
takes a conditional branch that the other doesn't, or when a computed
branch leads to different addresses in the two executions, or when $T$
ends, whichever comes first.
Past this point $H$ and $G$ diverge and are no longer directly
comparable, as they execute different instructions.
Our third requirement for being a witness is that each load in the
common period either must obtain the same value in $H$ and $G$, or
must itself be one of the $R'_i$ loads, or must be ordered in $H$
after one of the $R'_i$ loads (this is the remaining part of our
interpretation of ``all else being equal''.)

@@

For example, consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}''.
This would be a valid transformation, and the resulting behavior would
not count as OOTA according to our definition because the
dependencies in T0 and T1 would not be semantic.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and JITs on
one hand, and to link-time optimizations (LTO) on the other;
(2)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(3)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(4)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(5)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(6)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

>>>>

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.

We will use \emph{full C++} to denote the standard including this
prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
that excludes the prohibition but is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

In some cases, the standard permits architectures to differ as to whether
an execution contains a semantic dependency.
For example, consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
running on implementations in
which \co{char} is architecturally a signed type,
but not those for which it is unsigned.

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).\footnote{
	This requirement is the standard's ``as-if'' rule.}
This means:
(1)~The program's output must be the same as that of the abstract execution;
(2)~Volatile accesses ``are evaluated strictly according to the
rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
(3)~[A condition on the timing and interleaving of input and output,
which does not matter for our purposes].
We will say that the abstract execution is \emph{realized by} the
hardware execution.

@@

Conditional or computed branches also provide ordering.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional- or computed-branch
instruction must commit after the branch, and hence after the source
for the branch's condition and/or destination
input.\footnote{
	A branch that conditionally jumps to the
	next instruction would be an exception.
	We ignore this possibility by treating such branches as no-ops.}

@@

We begin by recognizing that semantic dependencies are relative to a
particular execution and a particular implementation,
as described in Section~\ref{sec:Properties of Semantic Dependencies}.
The same source code may or may not contain a semantic dependency,
according to the details of the execution in question and the
machine code produced by the compiler.
For this reason we will characterize semantic dependencies in a given
abstract execution realized by a given hardware execution.
(While it is possible to argue about semantic dependencies in abstract
executions that have no hardware realizations, doing so seems pointless.)

@@

Let $W'$ and $R'_i$ be the accesses in $H$ that $W$ and the
non-omitted $R_i$ loads map to.
We then require that the hardware executions of $T$ in $H$ and $G$ be
identical for an initial period lasting up to the first of the $R'_i$.
Following this initial period there will be a common period, during
which $H$ and $G$ execute the same machine instructions but do not
necessarily compute the same values.
This common period ends when one of the hardware executions
takes a conditional branch that the other doesn't, or when a computed
branch leads to different addresses in the two executions, or when $T$
ends, whichever comes first.
Past this point $H$ and $G$ diverge and are no longer directly
comparable, as they execute different instructions.
Our third requirement for being a witness is that each load in the
common period must either obtain the same value in $H$ and $G$, or
itself be one of the $R'_i$ loads, or be ordered in $H$
after one of the $R'_i$ loads (this is the remaining part of our
interpretation of ``all else being equal''.)

@@

For example, consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}'', as shown in
Section~\ref{sec:Global Optimization Can Destroy Dependencies}.
This would be a valid transformation, and the resulting behavior would
not count as OOTA according to our definition because the
dependencies in T0 and T1 would not be semantic.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and just-in-time
compilers (JITs) on
one hand, and to link-time optimization (LTO) on the other;
(2)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(3)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(4)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(5)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(6)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.
